---
title: "Notes 2024-09-19"
format: html
editor_options: 
  chunk_output_type: console
---

## Interpreting regression models



```{r}

library(tidyverse)

data.frame(x = c(0, 200), 
           y = c(-2000, 5000)) %>%
        ggplot(aes(x, y)) + 
        
        geom_blank() + # Draws an empty canvas with scales
        
        geom_abline(intercept = -1945, slope = 38.2) + 
        
        coord_cartesian(xlim = c(120, 180), 
                        ylim = c(1000, 5000))



```


Calculate VO2max when height is 155

```{r}

vo2max <- -1945 + 38.2 * 155

```



## Performance and VO2max


```{r}


intercept <- 1548
slope <- -12.4


data.frame(name = c("Anders", "Sigvard"), 
           VO2max = c(78.4, 72.9)) %>%
  mutate(performance = intercept + slope * VO2max) %>%
  
 
  
  ggplot(aes(VO2max, performance, color = name)) + 
  
  geom_abline(intercept = intercept, 
              slope = slope) +
  
  geom_point(size = 3)
  

```



## Dummy variables with an interaction 


```{r}
#| warning: false
#| message: false
#| echo: true


intercept <- 8.2
b1 <- 4.4
b2 <- -3.9
b3 <- 2.7

dat <- data.frame(group = c("ctrl", "ctrl", "expr", "expr"), 
                  sex = c("female", "male", "female", "male")) %>%
  print()

dat %>%
  # create a dummy variable
        mutate(xgroup = if_else(group == "ctrl", 0, 1), 
               xsex = if_else(sex == "female", 0, 1), 
               interaction = xgroup * xsex) %>%

  # do predictions
        mutate(delta_strength = intercept + b1 * xgroup + b2 * xsex + b3 * interaction)  %>% 
  
  ggplot(aes(group, delta_strength, fill = sex)) + 
  
  geom_bar(stat = "identity", position = position_dodge())
  
  



```




### Predict with error


```{r}

# Deterministic model

vo2max155 <- -1945 + 38.2 * 155
vo2max165 <- -1945 + 38.2 * 165


dat <- data.frame(pred = c(rnorm(1000, vo2max155, 72.2), 
                           rnorm(1000, vo2max165, 72.2)))



dat %>%
  mutate(x =c(rep(155, 1000), rep(165, 1000))) %>%
  ggplot(aes(x, pred)) + 
  
  
  coord_cartesian(xlim = c(145, 175), 
                  ylim = c(3500, 5000)) +
  
  geom_point(alpha = 0.1, color = "steelblue", size = 2) + 
  
  geom_abline(intercept = -1945, slope = 38.2)


```


### The hypertrophy data set

The hypertrophy data set can be used to set up hypotheses about e.g. what is associated with muscle growth. Muscle growth can be measured using vastus lateralis thickness. In the `hypertrophy` data set we have `VL_T1`, `VL_T2` etc containing information on vastus lateralis thickness in cm.


To create a model that gives us the average change in VL per group.

```{r}

library(exscidata)
library(tidyverse)

hypertrophy %>%
  print()

dat <- hypertrophy %>%
  select(PARTICIPANT, GROUP, VL_T1: VL_T4) %>%
  mutate(delta_vl = VL_T3 - VL_T1) %>%
  print()

factor(dat$GROUP)



m <- lm(delta_vl ~ GROUP, dat)

summary(m)


```


To get a model that tells us about the average change in VL when we use a continuous predictor in the data set. 


```{r}


dat <- hypertrophy %>%
  select(PARTICIPANT, GROUP, VL_T1: VL_T4, TESTOSTERONE_T1) %>%
  mutate(delta_vl = VL_T3 - VL_T1) %>%
  print()

m <- lm(delta_vl ~ TESTOSTERONE_T1, dat)

summary(m)



```





